Ownership types in theory and practice (in Rust)

### 2023-01-14 paper/paper.tex
Started work on adding memory safety example

>>>>

% TODO: clarify that reference and alias is used synonymously in this paper
\paragraph{Memory safety issue example}

% The following describes an example in which ownership types can prevent a memory security issue.

Ralf Jung \cite{understanding-evolving-rust} demonstrates the kind of memory safety issues that commonly arise in systems programming languages with the following example in C++.

\begin{lstlisting}
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points *into* v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

% TODO: use less "we"
First, we initialize a vector \emph{v} that contains two integers, which are stored in a buffer in memory.
% The initial contents of the vector \emph{vec} are stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
% TODO: change wording from source
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.

% This is possible because both pointers were aliasing: an action through a pointer (v) will in general also affect all its aliases (vptr).
More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).

% TODO: add graph from paper to visualize

% The fact that \emph{vptr} is now a dangling pointer becomes a problem in the last line, where we load from vptr.
Finally, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
% Such runtime bugs are very hard to track down.
% The main problem here is that such use-after-free bugs cannot be detected at compile time.

Now, a similar piece of code in Rust will explored.

\begin{lstlisting}
let mut v = vec![10, 11];
let vptr = &mut v[1]; // Points *into* ‘v‘.
v.push(12); // May reallocate the backing store of v.
println!("{}", *vptr); // Compiler error
\end{lstlisting}

% A quick definition of ownership: you cannot have mutability and alias at the same time.
% My own example:

% \begin{lstlisting}
% fn dangle() -> &String {
%   let s = String::from("hello world");
%   &s
% }
% \end{lstlisting}
